---
title: "Reproducible Research: Peer Assessment 1"
output: 
  html_document:
    keep_md: true
---


## Loading and preprocessing the data

```{r LoadData}
actData <- read.csv("activity.csv", stringsAsFactors=F)
stepsPerDay <- aggregate(steps ~ date, data=actData, FUN=sum)
```

```{r Histogram, echo=FALSE}
hh <- hist(stepsPerDay$steps, main="Histogram of steps per day", xlab="steps per day")
```

The histogram values are shown in the table below
```{r Table, echo=FALSE}
for (ii in 1:length(hh$counts)) {
  cat(sprintf("up to %5d: %3d\n", hh$breaks[ii+1], hh$counts[ii]))
}
```

## What is mean total number of steps taken per day?

```{r}
mean(stepsPerDay$steps, na.rm=T)
median(stepsPerDay$steps, na.rm=T)
``` 

## What is the average daily activity pattern?

```{r Intervals}
stepsPerInterval <- aggregate(steps ~ interval, data=actData, FUN=mean, na.rm=T)
```

```{r IntervalPlot, echo=FALSE}
plot(stepsPerInterval$steps, type='l', main="Mean number of steps per interval",
     xlab="Interval number (1-288)", ylab="mean number of steps")
```

Which interval has, on average across all the dates, 
the largest number of steps?

```{r MaxSteps}
which(stepsPerInterval$steps == max(stepsPerInterval$steps))
```

## Imputing missing values

A number of days have missing values:
```{r MissingValues}
table(is.na(actData$steps))
```

Fill with interval mean for same weekday

```{r FillData}
actData$Weekday <- factor(weekdays(as.Date(actData$date,format="%Y-%m-%d")), 
                          levels=c("Monday","Tuesday","Wednesday","Thursday",
                                   "Friday","Saturday","Sunday"), ordered=T)

dailyMean <- lapply(split(actData, actData$Weekday), 
                    function(d) aggregate(steps ~ interval, data=d, FUN=mean, na.rm=T))
actSplit <- split(actData, actData$Weekday)
for (wd in names(actSplit)) {
  actSplit[[wd]]$filled <- dailyMean[[wd]]$steps[match(actSplit[[wd]]$interval,dailyMean[[wd]]$interval)]
  actSplit[[wd]]$filled[!is.na(actSplit[[wd]]$steps)] <- actSplit[[wd]]$steps[!is.na(actSplit[[wd]]$steps)]
}
fillData <- unsplit(actSplit, actData$Weekday)
filledPerDay <- aggregate(filled ~ date, data=fillData, FUN=sum)
```

Having filled in missing values, the number of steps histogram is
```{r FilledHist, echo=FALSE}
hf <- hist(filledPerDay$filled, main="Filled steps per day", xlab="steps per day")
```

The corresponding mean and median are:

```{r FilledMeanMedian}
mean(filledPerDay$filled)
median(filledPerDay$filled)
```


## Are there differences in activity patterns between weekdays and weekends?

We classify observations into weekday and weekend dates.

```{r Weekends}
fillData$Weekend <- as.character(fillData$Weekday) %in% c("Saturday","Sunday")
fillData$dayType <- "weekday"
fillData$dayType[fillData$Weekend] <- "weekend"
fillData$dayType <- factor(fillData$dayType)
aggData <- tapply(fillData$filled, list(fillData$dayType, fillData$interval), mean)
```

The different patterns are:

```{r PlotWeekends, echo=FALSE}
par(mfrow=c(2,1))
plot(aggData[1,], type='l', main="Weekday", xlab="Interval", ylab="steps")
plot(aggData[2,], type='l', main="Weekend", xlab="Interval", ylab="steps")
```
