---
title: "RepData_PA1"
author: "Scott Cogswell"
date: "10/31/2016"
output: html_document
---

This report presents an analysis of activity monitoring data collected by a 
volunteer;
it should satisfy the requirement for the first peer assessment of the 
Reproducible Research course.

The objective is to answer the following questions:

* What is the mean number of steps taken each day
* What is the average daily activity pattern?

## Data

Our source data is from a personal activity monitoring device which collects 
the number of steps taken during every 5 minute interval throughout the day,
giving 288 observations per day.

The original dataset is 
[Activity monitoring data](https://d396qusza40orc.cloudfront.net/repdata%2Fdata%2Factivity.zip) [52K]
which includes the number of steps, the date, and the time interval.
The first step is to unzip the file, creating **activity.csv**.

## Methodology

The first step is to load the data and calculate the mean number of steps
per day.
```{r LoadData}
actData <- read.csv("activity.csv")
stepsPerDay <- aggregate(steps ~ date, data=actData, FUN=sum)
```

Eight days have no data whatsoever, these are ignored in this step.

```{r Histogram, echo=FALSE}
hh <- hist(stepsPerDay$steps, main="Histogram of steps per day", xlab="steps per day")
```

The histogram values are shown in the table below
```{r Table, echo=FALSE}
for (ii in 1:length(hh$counts)) {
  cat(sprintf("up to %5d: %3d\n", hh$breaks[ii+1], hh$counts[ii]))
}
```

The daily mean and median:
```{r}
mean(stepsPerDay$steps, na.rm=T)
median(stepsPerDay$steps, na.rm=T)
``` 
