---
title: "RepData_PA1"
author: "Scott Cogswell"
date: "10/31/2016"
output: html_document
---

This report presents an analysis of activity monitoring data collected by a 
volunteer;
it should satisfy the requirement for the first peer assessment of the 
Reproducible Research course.

The objective is to answer the following questions:

* What is the mean number of steps taken each day
* What is the average daily activity pattern?

We will consider each of these questions in turn.

## Data

Our source data is from a personal activity monitoring device which collects 
the number of steps taken during every 5 minute interval throughout the day,
giving 288 observations per day.

The original dataset is 
[Activity monitoring data](https://d396qusza40orc.cloudfront.net/repdata%2Fdata%2Factivity.zip) [52K]
which includes the number of steps, the date, and the time interval.
The first step is to unzip the file, creating **activity.csv**.

## Methodology

The first step is to load the data and calculate the mean number of steps
per day.
```{r LoadData}
actData <- read.csv("activity.csv", stringsAsFactors=F)
stepsPerDay <- aggregate(steps ~ date, data=actData, FUN=sum)
```

Eight days have no data whatsoever, these are ignored in this step.

### What is the mean number of steps taken each day?

```{r Histogram, echo=FALSE}
hh <- hist(stepsPerDay$steps, main="Histogram of steps per day", xlab="steps per day")
```

The histogram values are shown in the table below
```{r Table, echo=FALSE}
for (ii in 1:length(hh$counts)) {
  cat(sprintf("up to %5d: %3d\n", hh$breaks[ii+1], hh$counts[ii]))
}
```

The daily mean and median:
```{r}
mean(stepsPerDay$steps, na.rm=T)
median(stepsPerDay$steps, na.rm=T)
``` 

### What is the daily activity pattern?

To answer this question we aggregate by daily time interval.
```{r Intervals}
stepsPerInterval <- aggregate(steps ~ interval, data=actData, FUN=mean, na.rm=T)
```

```{r IntervalPlot, echo=FALSE}
plot(stepsPerInterval$steps, type='l', main="Mean number of steps per interval",
     xlab="Interval number (1-288)", ylab="mean number of steps")
```

Which interval has, on average across all the dates, 
the largest number of steps?

```{r MaxSteps}
which(stepsPerInterval$steps == max(stepsPerInterval$steps))
```

## Missing values

A number of days have missing values:
```{r MissingValues}
table(is.na(actData$steps))
```

There are 2,304 missing values in total. 
These are distributed according to days when our volunteer did not wear
the tracking device at all:
```{r MissingDays}
table(actData$date, is.na(actData$steps))
```

Is there a weekday bias from our non-observed days?

```{r Weekday}
actData$Weekday <- factor(weekdays(as.Date(actData$date,format="%Y-%m-%d")), 
                          levels=c("Monday","Tuesday","Wednesday","Thursday",
                                   "Friday","Saturday","Sunday"), ordered=T)
table(actData$Weekday)
table(actData$Weekday, is.na(actData$steps))
```

Our volunteer seems most diligent on Tuesday, having not missed one,
and least diligent on Fridays and Mondays, missing two of each.

To account for the missing values, we replace missing values with the mean
observed value for that same interval for the other weekdays for which we
have values. 

```{r FillData}
dailyMean <- lapply(split(actData, actData$Weekday), 
                    function(d) aggregate(steps ~ interval, data=d, FUN=mean, na.rm=T))
actSplit <- split(actData, actData$Weekday)
for (wd in names(actSplit)) {
  actSplit[[wd]]$filled <- dailyMean[[wd]]$steps[match(actSplit[[wd]]$interval,dailyMean[[wd]]$interval)]
  actSplit[[wd]]$filled[!is.na(actSplit[[wd]]$steps)] <- actSplit[[wd]]$steps[!is.na(actSplit[[wd]]$steps)]
}
fillData <- unsplit(actSplit, actData$Weekday)
filledPerDay <- aggregate(filled ~ date, data=fillData, FUN=sum)
```

Having filled in missing values, the number of steps histogram is
```{r FilledHist, echo=FALSE}
hf <- hist(filledPerDay$filled, main="Filled steps per day", xlab="steps per day")
```


The corresponding mean and median are:
```{r FilledMeanMedian}
mean(filledPerDay$filled)
median(filledPerDay$filled)
```

## Are there differences in activity patterns between weekdays and weekends?

We classify observations into weekday and weekend dates.

```{r Weekends}
fillData$Weekend <- as.character(fillData$Weekday) %in% c("Saturday","Sunday")
fillData$dayType <- "weekday"
fillData$dayType[fillData$Weekend] <- "weekend"
fillData$dayType <- factor(fillData$dayType)
aggData <- tapply(fillData$filled, list(fillData$dayType, fillData$interval), mean)
```

The different patterns are:

```{r PlotWeekends, echo=FALSE}
par(mfrow=c(2,1))
plot(aggData[1,], type='l', main="Weekday", xlab="Interval", ylab="steps")
plot(aggData[2,], type='l', main="Weekend", xlab="Interval", ylab="steps")
```



```